Further Evidence of Modified Spin-down in Sun-like Stars:\\Pileups in the Temperature--Period Distribution

### 2022-04-08 src/ms.tex
more real estate for Teff residuals

\edit1{An alternative explanation for the short-period overdensity is contamination by subgiants. Stars which have \teff on the ZAMS higher than $\sim$6250 will never undergo magnetic braking, spinning rapidly for the entirety of their main-sequence lifetimes. As these stars evolve away from the main-sequence, their temperatures drop while \logg changes minimally, while the spin rate slows due to radial expansion. Removing subgiants using spectroscopic parameters alone is difficult...}

@@

    \item The long-period pileup is well-described by a constant Rossby number, with a critical value of \rocritfinal, in the \teff range of $\approx$~5000-6250~K. The pileup is also populated by stars with a wide range of isochrone ages (\agerange). A pileup of stars with a constant Rossby number and a broad range of ages is a prediction of the WMB model of \citet{vanSaders2016, vanSaders2019}. The precise value of \rocrit is sensitive to \teff scale shifts between observational data and the models used to compute \taucz.

@@

    \item We \edit1{tentatively detect} a secondary overdensity of stars at the short-period edge of the \teff-\prot plane. This overdensity \edit1{appears to be} less prominent than the long-period overdensity, \edit1{possibly indicating that} the short-period pileup is shorter-lived ($\mathcal{O}(10^8 \text{yr})$) relative to the long-period pileup ($>10^9$~yr). The short-period pileup appears to be intermediate to the empirical Pleiades (0.12~Gyr) and Praesepe (0.67~Gyr) open cluster sequences and may result from a temporary epoch of stalled spin-down due to core–envelope coupling, as proposed by \curtis. The short-period pileup can also be fit with a constant Rossby model, though over a range of \teff that differs from that of the long-period pileup.

@@

% \begin{figure*}
%     \centering
%     \includegraphics[width=\linewidth]{./figures/ridge.pdf}
%     \caption{The \teff-\prot plane for the CKS sample. Point colors are scaled to the CKS ages determined from isochrone fitting. The source of \prot is denoted above each panel, where \citet{David2021} is a compilation of vetted periods, rather than a source of original measurements. The black trapezoid indicates the approximate area of the ridge. The grey curves indicate empirical cluster sequences from \curtis, corresponding to ages of $\sim$2.7, 1, 0.67, and 0.12~Gyr from top to bottom.}
%     \label{fig:ridge}
% \end{figure*}

@@

\edit1{We also note that the short-period pileup is not as prominent in the \lamostsan sample, compared to the \lamostmcq sample (see the rightmost panels of Figure~\ref{fig:xmatch}). The differences between the \mma and \santos catalogs may provide an explanation for this observation. The latter catalog contains a much larger number of detections at longer rotation periods, and the higher sensitivity to more slowly rotating stars within the \santos may then make the short-period pileup appear weaker relative to the long-period pileup. This is as expected since open cluster obsbervations suggest that the short-period pileup can not be a long-lived feature, while the long-period pileup appears to be much longer lived. Consequently, the \santos catalog may more accurately reflect the relative strengths of these two features.}

@@

\section{The Gaia Color–Period Plane}\label{app:gaia}
\edit1{Given that the long-period pileup was previously obscured by imprecise \teff measurements, we explored whether the feature could be recovered using the high-precision photometric colors provided by the Gaia mission \citep{Gaia2016}. We cross-matched the \mma and \santos samples with Gaia EDR3 \citep{GaiaEDR3} using a 1 arcsec search radius and the \texttt{astroquery} package \citep{astroquery}. We retrieved reddening estimates for each target by querying the \texttt{Bayestar19} 3D dust map using the \texttt{dustmaps} package \citep{dustmaps, Green2019}. We then compared the ($G_{BP}-G_{RP}$)–\prot distribution with constant Rossby curves. For this exercise, we used the empirically calibrated \taucz relation of \citet{Corsaro2021}, who presented \taucz as a quadratic function of Gaia $G_{BP}-G_{RP}$ color using the Kepler LEGACY asteroseismic sample as calibrators. We found that the long-period pileup is clearly visible in the ($G_{BP}-G_{RP}$)–\prot plane once stars with high reddening ($A_V>0.2$) are excluded. In order to match a curve of constant Rossby number, given by $\mathrm{Ro}=\mathrm{Ro_\odot}=0.496$ on the \citet{Corsaro2021} scale, we found that a -0.1 mag shift to the Gaia colors of the data was required (or, equivalently, a +0.1 mag shift applied to the constant Rossby curve). While there is no justification for such a large shift, it may indicate the presence of a systematic offset in the \taucz relation.}

>>>>

\edit1{Finally, we note that we regard the short-period pileup as less secure than the long-period pileup. In Appendix~\ref{app:harmonics} we show that the short-period pileup does not appear to be a harmonic of the long-period pileup, as one might expect if the periods on the short-period pileup were erroneously determined. However, the short-period pileup is primarily observed in the \lamostmcq sample, while it is unclear whether it is present in the \lamostsan sample (see the rightmost panels of Figure~\ref{fig:xmatch}). The differences between the \mma and \santos catalogs may provide an explanation for this observation. The latter catalog contains a much larger number of detections at longer rotation periods, and the higher sensitivity to more slowly rotating stars within the \santos may then make the short-period pileup appear weaker relative to the long-period pileup. This is as expected since open cluster observations suggest that the short-period pileup can not be a long-lived feature, while the long-period pileup appears to be much longer lived. Consequently, the \santos catalog may more accurately reflect the relative strengths of these two features.}

%An alternative explanation for the short-period overdensity is contamination by subgiants. Stars which have \teff on the ZAMS higher than $\sim$6250 will never undergo magnetic braking, spinning rapidly for the entirety of their main-sequence lifetimes. As these stars evolve away from the main-sequence, their temperatures drop while \logg changes minimally, while the spin rate slows due to radial expansion. While cursory attempts to isolate subgiants have been taken, it is

@@

    \item The long-period pileup is well-described by a constant Rossby number, with a critical value of \rocritfinal, in the \teff range of $\approx$~5500-6250~K. The pileup is also populated by stars with a wide range of isochrone ages (\agerange). A pileup of stars with a constant Rossby number and a broad range of ages is a prediction of the WMB model of \citet{vanSaders2016, vanSaders2019}. The precise value of \rocrit is sensitive to \teff scale shifts between observational data and the models used to compute \taucz.

@@

    \item We \edit1{tentatively detect} a secondary overdensity of stars at the short-period edge of the \teff-\prot plane. This overdensity \edit1{appears to be} less prominent than the long-period overdensity, \edit1{possibly indicating that} the short-period pileup is shorter-lived, $\mathcal{O}(10^8 \text{yr})$, relative to the long-period pileup ($>10^9$~yr). The short-period pileup appears to be intermediate to the empirical Pleiades (0.12~Gyr) and Praesepe (0.67~Gyr) open cluster sequences and may result from a temporary epoch of stalled spin-down due to core–envelope coupling, as proposed by \curtis. The short-period pileup can also be fit with a constant Rossby model, though over a range of \teff that differs from that of the long-period pileup.

@@

@@

@@

\section{The Gaia Color–Period Plane}\label{app:gaia}
\edit1{Given that the long-period pileup was previously obscured by imprecise \teff measurements, we explored whether the feature could be recovered using the high-precision photometric colors provided by the Gaia mission \citep{Gaia2016}. We cross-matched the \mma and \santos samples with Gaia EDR3 \citep{GaiaEDR3} using a 1 arcsec search radius and the \texttt{astroquery} package \citep{astroquery}. We retrieved reddening estimates for each target by querying the \texttt{Bayestar19} 3D dust map using the \texttt{dustmaps} package \citep{dustmaps, Green2019}. We then compared the ($G_{BP}-G_{RP}$)–\prot distribution with constant Rossby curves. For this exercise, we used the empirically calibrated \taucz relation of \citet{Corsaro2021}, who presented \taucz as a quadratic function of Gaia $G_{BP}-G_{RP}$ color using the Kepler LEGACY asteroseismic sample as calibrators. We found that the long-period pileup is clearly visible in the ($G_{BP}-G_{RP}$)–\prot plane once stars with high reddening ($A_V>0.2$) are excluded. In order to match a curve of constant Rossby number, given by $\mathrm{Ro}=\mathrm{Ro_\odot}=0.496$ on the \citet{Corsaro2021} scale, we found that a $\approx$-0.1 mag shift to the Gaia colors of the data was required (or, equivalently, a +0.1 mag shift applied to the constant Rossby curve). While there is no justification for such a large shift, it may indicate the presence of a systematic offset in the \taucz relation. We also note that we do not have an explanation for why the solar value Rossby curve does not pass through the Sun.}
